% 4
\chapter{ARM Hardware}


\begin{node}\label{arm-0000}%
These are just some reading notes as I work my way through the ARM
specification. The ARMv8-M are for ``microcontroller profiles'', the
ARMv8-A are for ``application profiles'' (i.e., the usual desktop
computers), the ARMv8-R are for ``real-time profiles''. The Arm
reference manual gives the specification for the behaviour of an
abstract machine called the \define{Processing Element} (or PE for
short).

As far as formalization, Anthony Fox~\cite{fox2001arm} formalized the ARM6
specification using HOL. Fox~\cite{fox2002arm} summarizes the
formalization using techniques found in earlier reports~\cite{fox2001algebraic}.
The trick was to use logical relations to
encode circuits --- so the NAND gate is an ``atomic'' predicate
$NAND[i_{1},i_{2},o]$ which satisfies
$NAND[i_{1},i_{2},o]=(o\iff\neg(i_{1}\land i_{2}))$. We can feed the
input of one circuit $C_{1}[i,o]$ as input into another $C_{2}[i,o]$ and
this is the equation $C_{12}[i,o]=(\exists x)(C_{1}[i,x]\land C_{2}[x,o])$.
\end{node}

\begin{node}[Definitions]\label{arm-0001}%
Regarding cold, warm, and hard boots.\footnote{From \url{https://www.lenovo.com/us/en/glossary/cold-boot/}}

\begin{node}\label{arm-0002}%
A \define{Cold Boot} involves starting the computer from a completely
off state. The basic steps involved in a cold boot:
\begin{enumerate}
\item The computer's power supply turns on and supplies electricity to
  the system;
\item The CPU initializes the system's firmware UEFI (unified extensible
  firmware interface) or for older machines BIOS (basic
  input/output system) --- I refer to ``BIOS'' the same way I use
  ``Kleenex'' instead of ``tissue paper'';
\item BIOS performs the POST (power-on self-test) which checks the
  hardware components, and ensures they are functioning correctly;
\item\label{arm:last-step-of-cold-boot} BIOS then searches and loads the operating system from the
  computer's storage.
\end{enumerate}
\end{node}

\begin{node}\label{arm-0003}%
A \define{Warm Boot} refers to restarting the computer without powering
it completely off. During a warm boot, it skips to
step~\ref{arm:last-step-of-cold-boot} of the cold boot process.
\end{node}
\end{node} % Definitions

