% 51
\section{Basic Definitions}

\begin{node}[Size issues]\label{cat-0002}%
Depending on our choice of set theory, we have different notions of
``small'' and ``large'' collections. In \NBG\ set theory~\zref{sec:set:nbg}
``large'' collections are proper classes, and ``small'' collections are
sets. In Tarski--Grothendieck set theory~\zref{sec:set:tarski-grothendieck},
we usually pick some ``smallest'' Grothendieck universe $\mathcal{U}$
containing $\NN$, which would also contain $\RR$, $\CC$, and all other
sets we use in classical mathematics, then consider elements of
$\mathcal{U}$ as small collections. Since there's possibly more than one
choice of universe, we speak of ``$\mathcal{U}$-small'' and
``$\mathcal{U}$-large'' sets depending on if they are elements of
$\mathcal{U}$ or not. Further, subsets of $\mathcal{U}$ are sometimes
called ``$\mathcal{U}$-moderate''. For a review of the different notions
of ``small'' and ``large'' collections germane to category theory, see
Shulman~\cite{shulman2008set}.

For various constructions in Category theory, \NBG\ will prove not
enough. We could try \MK\ axioms, but it will still not be enough. Some
authors invent a notion of ``conglomerates'' for forming the collection
of all proper classes. But this will turn out to still not be enough. This
is why we turn to Tarski--Grothendieck set theory.
\end{node}

\begin{node}[Category]\label{cat-0001}%
\begin{definition}[Version 1]
A \define{Category} $\cat{C}$ consists of
\begin{enumerate}
\item a collection of objects $\ob(\cat{C})$;
\item for any objects $X,Y\in\ob(\cat{C})$, there is a (possibly empty)
  collection of morphisms $\hom(X,Y)$;
\item for each object $X\in\ob(\cat{C})$, an identity morphism $\id_{X}\in\hom(X,X)$;
\item for objects $X,Y,Z\in\ob(\cat{C})$, there is a binary operator
  $\circ\colon\hom(Y,Z)\times\hom(X,Y)\to\hom(X,Z)$ called
  \define{composition} (of morphisms);
  % and morphisms $f\in\hom(X,Y)$, $g\in\hom(Y,Z)$, there exists a morphism $g\circ f\in\hom(X,Z)$
\end{enumerate}
such that
\begin{enumerate}
\item composition is associative: for any $W$, $X$, $Y$, $Z\in\ob(\cat{C})$, if
  $f\in\hom(W,X)$, $g\in\hom(X,Y)$, $h\in\hom(Y,Z)$, then $(h\circ g)\circ f=h\circ(g\circ f)$;
\item unit laws for composition: for any objects $X$, $Y\in\ob(\cat{C})$,
  if $f\in\hom(X,Y)$, then $\id_{Y}\circ f=f$ and $f\circ\id_{X}=f$.
\end{enumerate}
\end{definition}

\begin{definition}[Version 2]\label{cat-0000}%
A \define{Category} $\cat{C}$ consists of
\begin{enumerate}
\item a collection of objects $\ob(\cat{C})$;
\item a collection of morphisms $\hom(\cat{C})$
\end{enumerate}
equipped with
\begin{enumerate}
\item a mapping $\id\colon\ob(\cat{C})\to\hom(\cat{C})$ assigning to
  each object its identity morphism;
\item a mapping $\circ\colon\hom(\cat{C})\times\hom(\cat{C})\to\hom(\cat{C})$;
\item a source and target mapping $s,t\colon\hom(\cat{C})\to\ob(\cat{C})$;
\end{enumerate}
such that
\begin{enumerate}
\item source and target are respected by composition: $s(g\circ f)=s(f)$
  and $t(g\circ f)=t(g)$;
\item source and target are respected by identities: $s(\id_{X})=X$ and $t(\id_{X})=X$;
\item composition is associative: $(h\circ g)\circ f=h\circ(g\circ f)$
  whenever $t(f)=s(g)$ and $t(g)=s(h)$;
\item composition satisfies the unit laws: if $s(f)=X$ and $t(f)=Y$,
  then $\id_{Y}\circ f=f$ and $f\circ\id_{X}=f$.
\end{enumerate}
\end{definition}

\begin{node}[Examples]\label{cat-0007}%
\begin{node}\label{cat-0004}%
The large category $\Set$ of sets\dots depends greatly on your choice of
foundations. ``Morally'', $\ob(\Set)$ consists of all [``small''] sets, and its
morphisms are [``small''] functions between those sets. But the exact
details varies depending on whether our foundations is predicative or
impredicative, classical or intuitionistic, admit or negate the axiom of
choice, and so on.
\end{node}

\begin{node}\label{cat-000A}%
If $\mathcal{U}$ is a Grothendieck universe~\zref{tg-0001}, then we can
form $\mathcal{U}\mbox{-}\Set$ whose collection of objects is precisely
$\mathcal{U}$ and whose collection of morphisms is precisely the
functions which are elements of $\mathcal{U}$.
\end{node}

\begin{node}\label{cat-0005}%
The large category $\cat{Ens}$ or $\cat{SET}$ is the category whose object is
the proper class of all sets $\ob(\cat{SET})=\ob(\cat{Ens})=V$ and whose
morphisms is all functions. This is a large category.
\end{node}

\begin{node}\label{cat-0003}%
The large category $\Grp$ of groups consists of $\ob(\Grp)$ the collection of all
``small'' groups [i.e., whose underlying set belongs to the universe of
sets], and the morphisms are all group morphisms.
\end{node}

\begin{node}\label{cat-0006}%
The large category $\Top$ of topological spaces consists of the collection of
all ``small'' topological spaces and morphisms are continuous maps
between them.
\end{node}

\begin{node}\label{cat-0008}%
Let $\FF$ be a field. The large category $\Vect{\FF}$ has its objects be
[``small''] vector spaces over $\FF$, and morphisms are $\FF$-linear
mappings between them.
\end{node}

\begin{node}\label{cat-000F}%
The category $\Mfld$ of $C^{\infty}$ manifolds is a category whose
objects are smooth manifolds, and morphisms are smooth mappings between
them. 
\end{node}

\begin{node}\label{cat-000M}%
Let $\cat{C}$ be a category consisting of a single object. This encodes
a monoid in the collection of morphisms.
\end{node}
\end{node} % examples

\begin{definition}\label{cat-000O}%
Let $\cat{C}$ be a category. We may form the \define{Opposite} category
of $\cat{C}$, denoted $\cat{C}^{op}$, which consists of the same objects
but a morphism $f\colon X\to Y$ in $\cat{C}^{op}$ is precisely the same
as a morphism $f\colon Y\to X$ in $\cat{C}$, and composition of
morphisms in the opposite category $g\circ f$ is precisely the reverse
of composition in the original category $(f\circ g)_{orig}$.

\begin{node}[Duality principle]\label{cat-000P}%
Every construction we can do in a category $\cat{C}$ has a dual
construction in the opposite category $\cat{C}^{op}$ which is usually
prefixed by ``co-''.
\end{node}
\end{definition}

\begin{definition}\label{cat-000G}%
Let $\cat{C}$ be a category. We define a \define{Subcategory} of $\cat{C}$
to consist of a category $\cat{D}$ such that $\ob(\cat{D})\subset\ob(\cat{C})$
and for all objects $X,Y\in\ob(\cat{D})$,
$\hom_{\cat{D}}(X,Y)\subset\hom_{\cat{C}}(X,Y)$. The composition of
morphisms in $\cat{D}$ is given by restricting the composition operator
from $\cat{C}$. The identity morphisms in $\cat{D}$ are precisely the
identity morphisms of $\cat{C}$ restricted to the objects of $\cat{D}$.

\begin{definition}\label{cat-000H}%
Let $\cat{D}$ be a subcategory of $\cat{C}$. We say $\cat{D}$ is a
\define{Full} subcategory if for all objects $X,Y\in\ob(\cat{D})$, we
have $\hom_{\cat{D}}(X,Y)=\hom_{\cat{C}}(X,Y)$.
\end{definition}
\end{definition}

\begin{node}[Types of Objects]\label{cat-0012}%

\begin{definition}\label{cat-0013}%
In a category $\cat{C}$, a \define{Terminal Object} is an object usually
denoted $\mathbf{1}\in\ob(\cat{C})$ such that for any object
$X\in\ob(\cat{C})$ there exists exactly one morphism $!\colon X\to\mathbf{1}$.
\end{definition}

\begin{node}[Examples]\label{cat-0014}%
\begin{enumerate}
\item In $\Set$, any singleton set is a terminal object.
\item In $\Grp$, the trivial group is a terminal object.
\end{enumerate}
\end{node}

\begin{definition}\label{cat-0015}%
Let $\cat{C}$ be a category. We define an \define{Initial Object} in
$\cat{C}$ to be an object $\emptyset\in\cat{C}$ (or in older literature $\mathbf{0}\in\cat{C}$)
such that for each object $X\in\cat{C}$ there exists a unique morphisms
$\emptyset\to X$.
\end{definition}

\begin{node}[Examples]\label{cat-0016}%
\begin{enumerate}
\item In $\Set$, the empty set is the initial object.
\item In $\Grp$, the trivial group is an initial object.
\end{enumerate}
\end{node}

\begin{definition}\label{cat-0017}%
In a category $\cat{C}$, we define a \define{Null Object} (or
\textit{Zero Object}) to be an
object $\mathbf{0}\in\cat{C}$ (or $Z\in\cat{C}$) which is both initial
and terminal.
\end{definition}
\end{node}

\begin{node}[Types of morphisms]\label{cat-000W}%

\begin{definition}\label{cat-000X}%
Let $\cat{C}$ be a category. We call a morphism $f\colon A\to B$ in
$\cat{C}$ an \define{Isomorphism} if there exists a morphism $g\colon B\to A$
in $\cat{C}$ such that $f\circ g=\id_{B}$ and $g\circ f=\id_{A}$.
In this case, $g$ is unique, and we denote it by $f^{-1}$, and further
we say $A$ and $B$ \emph{are isomorphic}. We indicate this by writing
$A\iso B$ to indicate they are isomorphic objects (i.e., there exists an
isomorphism from $A$ to $B$).
\end{definition}

\begin{definition}\label{cat-000Y}%
Let $\cat{C}$ be a category.
We call a morphism $m\colon B\to C$ \define{Monic} (or a \textit{Monomorphism})
if for all morphisms $f_{1},f_{2}\colon A\to B$ such that $m\circ f_{1}=m\circ f_{2}$
implies $f_{1}=f_{2}$.
\end{definition}

\begin{node}[Example]\label{cat-000Z}%
In $\Set$ (resp., $\Grp$), monomorphisms are precisely injective functions
(resp., group morphisms). The mnemonic to remember is that
``Monomorphisms go \textit{mono-e-mono}''.
\end{node}

\begin{definition}\label{cat-0010}%
Let $\cat{C}$ be a category.
We call a morphism $e\colon B\to C$ \define{Epic} (or a \textit{Epimorphism})
if for all morphisms $g_{1},g_{2}\colon C\to D$ in $\cat{D}$, 
whenever $g_{1}\circ e=g_{2}\circ e$ then we necessarily have $g_{1}=g_{2}$.
\end{definition}

\begin{node}[Example]\label{cat-0011}%
In $\Set$ (resp., $\Grp$), epimorphisms are precisely surjective functions
(resp., group morphisms).
\end{node}
\end{node}
\end{node} % category

\begin{node}[Groupoids]\label{cat-000B}%
\begin{definition}\label{cat-000C}%
A \define{Groupoid} is a category $\cat{G}$ whose morphisms are all
isomorphisms. It is common to call $\ob(\cat{G})$ the \textit{base} of
the groupoid.
\end{definition}

\begin{node}[Examples]\label{cat-000D}%
\begin{node}\label{cat-000I}%
Let $G$ be a group. Then we can form a groupoid whose set of objects is
just a singleton set, and collection of morphisms is $G$, and
composition operator is the law of composition from the group structure
on $G$. We can reverse this intuition, and assert \emph{all} groups are
``just'' one-object groupoids.
\end{node}

\begin{node}[Fundamental groupoid]\label{cat-000E}%
For any topological space $X$, we may form the fundamental groupoid
$\Pi_{1}(X)$ whose objects are the set of underlying points of $X$ and
morphisms $f\colon A\to B$ are homotopy equivalence classes of paths
from $A$ to $B$.  Composition then concatenates paths.
\end{node}

\begin{node}\label{cat-000Q}%
One rich vein for groupoids is describing the symmetry of fibred objects.
\end{node}

\begin{node}\label{cat-000R}%
Let $R$ be an equivalence relation on a set $X$. Then we can form a
groupoid $\cat{G}$ whose set of objects are $\ob(\cat{G})=X$ and $R$
gives us the non-identity morphisms: there is a $f\colon x\to y$ iff
$R(x,y)$ holds. We then interpret $R(y,x)$ as $f^{-1}\colon y\to x$. The
resulting category $\cat{G}$ is indeed a groupoid.
\end{node}
\end{node}
\end{node}

\begin{node}[Functors]\label{cat-0009}%
When we define a new mathematical gadget, it will be useful to turn our
attention to defining morphisms for our new gadget. We call the
morphisms of categories ``functors''.

\begin{definition}
Let $\cat{C}$, $\cat{D}$ be categories. We define a \define{Functor}
from $\cat{C}$ to $\cat{D}$ to be $F\colon\cat{C}\to\cat{D}$ consisting
of
\begin{enumerate}
\item a mapping of objects $F_{ob}\colon\ob(\cat{C})\to\ob(\cat{D})$;
\item a mapping of morphisms $F_{mor}\colon\hom_{\cat{C}}(X,Y)\to\hom_{\cat{D}}(F_{ob}(X),F_{ob}(Y))$;
\end{enumerate}
such that
\begin{enumerate}
\item $F$ preserves composition: $F_{mor}(g\circ f)=F_{mor}(g)\circ F_{mor}(f)$ whenever
  $g\circ f$ is defined in $\cat{C}$;
\item $F$ preserves identity morphisms: for each object $X\in\cat{C}$, $F_{mor}(\id_{X})=\id_{F_{ob}(X)}$.
\end{enumerate}
We often suppress the subscripts and just write things like
\[F\left(X\xrightarrow{f}Y\right)\quad=\quad F(X)\xrightarrow{F(f)}F(Y).\]
\end{definition}

\begin{node}[Examples]\label{cat-000J}%
\begin{node}\label{cat-000L}%
Let $\cat{C}$ be a category. We define the \define{Identity functor}
$\Id_{\cat{C}}$ which is just the identity function on objects, and the
identity function on morphisms.
\end{node}
  
\begin{node}[Linear representations of groups]\label{cat-000K}%
Let $\FF$ be a field, let $G$ be a group. We know we can form the
groupoid $\cat{G}$ from $G$ \zref{cat-000I}. Then $\FF$-linear
representations of $G$ are precisely the functions $\cat{G}\to\Vect{\FF}$.
\end{node}

\begin{node}\label{cat-000N}%
Let $\cat{C}$ be a locally-small category, let $X,Y\in\ob(\cat{C})$ be
objects. Then $\hom(X,-)\colon\cat{C}\to\Set$ maps each object
$B\in\cat{C}$ to the set of morphisms $\hom(X,A)$ and each function
$f\colon A\to B$ to $\hom(X,f)\colon\hom(X,A)\to\hom(X,B)$ sending
$g\mapsto f\circ g$ for each $g\in\hom(X,A)$.

We also have $\hom(-,Y)\colon\cat{C}^{op}\to\Set$ be a functor sending each
object $A\in\ob(\cat{C})$ to the set of morphisms $\hom(A,Y)$, and each
morphism $h\colon A\to B$ in $\cat{C}$ to $\hom(h,Y)\colon\hom(B,Y)\to\hom(A,Y)$
sending $g\in\hom(B,Y)$ to $g\circ h\in\hom(A,Y)$.
\end{node}
\end{node} % examples

\begin{node}[Types of functors]\label{cat-001A}%

\begin{definition}\label{cat-001D}%
We call a functor $F\colon\cat{C}\to\cat{D}$ an \define{Isomorphism}
if there exists a functor $G\colon\cat{D}\to\cat{C}$ such that $F\circ G=\Id_{\cat{D}}$
and $G\circ F=\Id_{\cat{C}}$. In this case, we write $G=F^{-1}$ and say
$\cat{C}$ and $\cat{D}$ are isomorphic.

(This coincides with the criteria~\zref{cat-000X} for a morphism in a
category to be an isomorphism.)
\end{definition}

\begin{definition}\label{cat-0019}%
We call a functor $F\colon\cat{C}\to\cat{D}$ \define{Faithful} if the
mapping of $\hom$-sets $F\colon\hom_{\cat{C}}(X,Y)\to\hom_{\cat{D}}(F(X),F(Y))$
is injective.
\end{definition}

\begin{definition}\label{cat-0018}%
We call a functor $F\colon\cat{C}\to\cat{D}$ \define{Full} if the
mapping of $\hom$-sets $F\colon\hom_{\cat{C}}(X,Y)\to\hom_{\cat{D}}(F(X),F(Y))$
is surjective.
\end{definition}

\begin{definition}\label{cat-001E}%
We call a functor $F\colon\cat{C}\to\cat{D}$ \define{Essentially Surjective} if
for each object $Y\in\cat{D}$ there exists an object $X\in\cat{C}$ such
that $F(X)\iso Y$ are isomorphic in $\cat{D}$.
\end{definition}

\end{node} % types of functors

\end{node}

\begin{node}[Natural transformations]\label{cat-000S}%

\begin{definition}
Let $F,G\colon\cat{C}\to\cat{D}$ be functors. A \define{Natural Transformation}
$\alpha\colon F\natTo G$ is a function which assigns to each object
$C\in\cat{C}$ an arrow $\alpha_{C}\colon F(C)\to G(C)$ such that for all
morphisms $f\colon C\to D$ in $\cat{C}$ the following diagram commutes
\begin{equation*}
\begin{tikzcd}[sep=large]
  C\ar[d,swap,"f"]\\
  D
\end{tikzcd}
\qquad
\begin{tikzcd}[column sep=4em,
    execute at end picture={%
 \node[blue,draw,very thick,fit={(\tikzcdmatrixname.north west) ([xshift=1ex]\tikzcdmatrixname-1-1.east|-\tikzcdmatrixname.south)},
    inner sep=1ex,label={[blue]above:$F(\cat{C})$}]{};
 \node[red,draw,very thick,fit={(\tikzcdmatrixname.north east) ([xshift=-1ex]\tikzcdmatrixname-1-2.west|-\tikzcdmatrixname.south)},
 inner sep=1ex,label={[red]above:$G(\cat{C})$}]{};
    }]%[sep=large]
F(C)\ar[d,swap,"F(f)"]\ar[r,"\alpha_{C}"] & G(C)\ar[d,"G(f)"]\\
F(D)\ar[r,swap,"\alpha_{D}"] & G(D)
\end{tikzcd}
\end{equation*}
\end{definition}

\begin{node}[Whiskering]\label{cat-000T}%
Let $F,G\colon\cat{C}\to\cat{D}$ be functors and $\alpha\colon F\natTo G$
be a natural transformation. If $H\colon\cat{D}\to\cat{E}$ is another
functor, we can form a natural transformation denoted $H\alpha\colon H\circ F\natTo H\circ G$
defined by $(H\alpha)_{C}=H(\alpha_{C})$ for each $C\in\ob(\cat{C})$.

Similarly, if $K\colon\cat{B}\to\cat{C}$ is a functor, we can define
another natural transformation $\alpha K\colon F\circ K\natTo H\circ K$
defined by $(\alpha K)_{B}=\alpha_{K(B)}$ for each $B\in\ob(\cat{B})$.
\end{node}

\begin{node}[Examples]\label{cat-000U}%

\begin{node}\label{cat-000V}%
Let $F\colon\cat{C}\to\cat{D}$ be a functor. We have the identity
natural transformation $\Id_{F}\colon F\natTo F$ defined by components
$(\Id_{F})_{C}=\id_{F(C)}$ for each object $C\in\ob(\cat{C})$.
\end{node}
\end{node}

\begin{definition}\label{cat-001B}%
Let $F,G\colon\cat{C}\to\cat{D}$ be functors. We define a
\define{Natural Isomorphism} to be a natural transformation
$\alpha\colon F\natTo G$ such that every component $\alpha_{X}$ (for
each $X\in\ob(\cat{C})$) is an isomorphism.
\end{definition}

\begin{definition}\label{cat-001C}%
A functor $F\colon\cat{C}\to\cat{D}$ is an \define{Equivalence of Categories}
(and we say $\cat{C}$ and $\cat{D}$ are \textit{equivalent}) if there
exists a functor $G\colon\cat{D}\to\cat{C}$ and natural isomorphisms
$F\circ G\iso\Id_{\cat{D}}\colon\cat{D}\to\cat{D}$ and
$\Id_{\cat{C}}\iso G\circ F\colon\cat{C}\to\cat{C}$. (The order of the
isomorphisms allows for a generalization later to adjoint functors.)
This is a weaker notion than isomorphism of categories~\zref{cat-001D}.
\end{definition}

\end{node}

\section{Universal Properties}
\input{tex/cat/universal-prop}
\section{Internalization}
\input{tex/cat/internal}
