% 20

\begin{node}\label{hda-0000}%
The basic idea of ``higher-dimensional algebra'' appears to be to
consider ways to define ``$n$-gadgets'' where $1$-gadgets are the
familiar things we learn in undergraduate abstract algebra (groups,
rings, vector spaces, etc.), $2$-gadgets are the internalized versions
in $\cat{Cat}$. This leads to consider different ways to categorify
these notions, and considerations of different ways to form
$n$-categories and internalization in there. This can also lead to
intriguing notions (like a topological category, being a category object
internal to $\Top$ the category of topological spaces).
\end{node}

\begin{node}[Lie 2-algebras]\label{hda-0001}%
Baez and Crans~\cite{baez2010higherdimensional} are a good reference for
Lie 2-algebras with an eye towards working analogously towards Lie
2-groups in previous work~\cite{baez2004higherdimensional}. Pradines'
review~\cite{pradines2007ehresmanns} may be worth perusing as well.
\end{node}

\begin{definition}\label{hda-0002}%
Let $K$ be a commutative ring.
A \define{$K$-Linear Category} (or \define{$K$-Algebroid}) is a category
enriched over $\Vect{K}$, i.e., whose
$\hom$-sets are vector spaces and whose composition operator is bilinear.
(This is a horizontal categorification of the notion of an associative
algebra; therefore, a Lie algebroid does not fit into this definition.) 
\end{definition}

\begin{node}[Lie Algebra objects]\label{hda-0005}%
\begin{definition}\label{hda-0004}%
Let $K$ be a commutative unital ring, let us consider a symmetric monoidal
$K$-linear category $(\cat{C},\otimes,\mathbf{1})$ with braiding $\tau$.
We define a \define{Lie algebra object} in $(\cat{C},\otimes,\mathbf{1},\tau)$
to consist of an object $L\in\cat{C}$ equipped with the Lie bracket
morphism $[-,-]\colon L\otimes L\to L$ such that
\begin{enumerate}
\item the Jacobi identity holds: $\left[-,\left[-,-\right]\right]
+ \left[-,\left[-,-\right]\right]\circ(\id_{L}\otimes\tau_{L,L})\circ(\tau\otimes \id_{L})
+ \left[-,\left[-,-\right]\right] \circ (\tau_{L,L}\otimes \id_{L})\circ (\id_{L}\otimes\tau_{L,L}) = 0$;
and
\item Antisymmetry: $[-,-]+[-,-]\circ \tau_{L,L}=0$.
\end{enumerate}
\end{definition}
\begin{node}[Example: recover usual notion]\label{hda-0006}%
We can recover the usual notion of a Lie algebra by taking $K$ to be a
field, $\cat{C}=\Vect{K}$ the category of vector spaces over $K$
equipped with the usual notion of the tensor product of vector
spaces. Then a Lie algebra object in $\cat{C}$ is just the usual notion
of a Lie algebra.
\end{node}
\begin{node}[Example: Super Lie algebras]\label{hda-0007}%
Let $K$ be a field and $\cat{C}=\sVect{K}$ be the category of super
vector spaces over $K$ equipped with the super-tensor product. Then a
Lie algebra object in $\cat{C}$ is just a super Lie algebra as we know
and love.
\end{node}
\begin{node}[Generalizations]\label{hda-0008}%
We can think of $L_{\infty}$-algebras as Lie algebras whose identities
hold up to homotopic equivalence. From this perspective,
$L_{\infty}$-algebras are a (vertical) categorification of Lie algebras.
\end{node}
\end{node}

\begin{definition}\label{hda-0009}%
A \define{Lie groupoid} is a groupoid internal~\zref{internal-0002} to the
category of smooth manifolds $\Mfld$. The ``infinitesimal
approximation'' to a Lie groupoid is precisely a Lie
algebroid~\zref{hda-0003}.

In particular, the collection of objects $G_{0}$ and morphisms $G_{1}$
are smooth manifolds; the source and target morphisms $G_{1}\to G_{0}$ are
surjective submersions; the identity assignment $\id$, the inverse map
$(-)^{-1}\colon G_{1}\to G_{1}$, and the composition operator are all
smooth mappings.

\begin{node}[Remark]\label{hda-000A}%
Usually we speak of a Lie groupoid over a manifold $M$, which considers
a Lie groupoid whose smooth manifold of objects is $M$.
\end{node}

\begin{node}[History]\label{hda-000B}%
Ehrasmann invented the notion of Lie groupoids as an alternative to the
framework of principal fibre bundles, specifically to study higher
prolongations of vector fields (and other related differential gadgets).
\end{node}
\end{definition}

\begin{node}[Lie Algebroid]\label{hda-000C}%
\begin{definition}\label{hda-0003}%
Let $X$ be a manifold. A \define{Lie Algebroid} over $X$ consists of
a vector bundle $E\to X$ equipped with
\begin{enumerate}
\item a Lie bracket $[-,-]\colon\Gamma(E)\otimes\Gamma(E)\to\Gamma(E)$
  on the space of sections;
\item a morphism of vector bundles (sometimes called the \textit{Anchor})
  $\rho\colon E\to TX$ whose tangent map preserves the bracket:
  $\D\rho([\xi,\zeta]_{\Gamma E})=[\D\rho(\xi),\D\rho(\zeta)]_{\Gamma TX}$
\end{enumerate}
such that the Leibniz rule holds: for all $X$, $Y\in\Gamma(E)$, and for
all $f\in C^{\infty}(E)$, we have $[X,f\cdot Y]=f\cdot[X,Y]+\rho(X)(f)\cdot Y$.
\end{definition}

\begin{node}[Examples]\label{hda-000D}%

\begin{node}\label{hda-000E}%
A Lie algebra is a Lie algebroid over a singleton $X=\{*\}$.
\end{node}

\begin{node}\label{hda-000F}%
If the anchor $\rho\colon E\to TX$ is trivial, then a Lie algebroid is a
bundle of Lie algebras.
\end{node}
\end{node}

\begin{node}[Further reading]\label{hda-000G}%
Fernandes and Crainic's lecture notes~\cite{fernandes2006lectures}
is worth reading, but there's a lot of literature on Lie algebroids (and
Lie groupoids).
\end{node}
\end{node}

\begin{node}[Stacks]\label{hda-000H}%
We can categorify the notion of a sheaf, and this is precisely a
stack. There are a number of subtleties surrounding ``correct''
definitions here, see Metzler~\cite{metzler2003topological} for details.
\end{node}

\begin{node}\label{hda-000I}%
The dg-category of chain complexes plays the role of stable
$(\infty,1)$-category of ``$(infty,1)$-vector spaces''. It's in this
sense that $L_{\infty}$-algebras are categorified Lie algebras. For more
about stable $\infty$-categories, see Lurie~\cite{lurie2009stable}.
\end{node}

\begin{node}\label{hda-000J}%
In general, for categorification of gadgets useful in representation
theory, see Savage~\cite{savage2015introduction}.
\end{node}