% 4
\begin{node}[Syntax]\label{prop-000J}%
\begin{node}\label{prop-000I}%
The BNF grammar for propositional logic may be written down explicitly
as a half-dozen production rules with variables treated as a
``primitive''. These rules tell us that variables are propositions,
so are negations of propositions, conjunction, disjunction, implication,
and bi-implication of propositions form propositions (respectively).
\begin{center}
\begin{tabular}{rcl}
$\langle$\textit{proposition}$\rangle$ & ::= & $\langle$\textit{variable}$\rangle$\\
& $|$ & $\neg\langle$\textit{proposition}$\rangle$\\
& $|$ & $\langle$\textit{proposition}$\rangle\land\langle$\textit{proposition}$\rangle$\\
& $|$ & $\langle$\textit{proposition}$\rangle\lor\langle$\textit{proposition}$\rangle$\\
& $|$ & $\langle$\textit{proposition}$\rangle\implies\langle$\textit{proposition}$\rangle$\\
& $|$ & $\langle$\textit{proposition}$\rangle\iff\langle$\textit{proposition}$\rangle$\\
\end{tabular}
\end{center}
\end{node}

\begin{definition}\label{prop-0002}%
Let $V$ be a (nonempty, countably infinite) set of propositional variables. Then
we inductively define a \define{Proposition} according to the following
rules:
\begin{enumerate}
\item Any propositional variable $A\in V$ is a proposition;
\item If $A$ and $B$ are propositions, then so are $\neg A$, $A\land B$,
  $A\lor B$, $A\implies B$, and $A\iff B$.
\end{enumerate}
Only those expressions constructed by these rules are propositions.

\begin{node}\label{prop-0003}%
We usually keep the set of variables $V$ implicit throughout. So we will
not explicitly refer to it unless absolutely necessary. We will
understand that ``a proposition'' really means ``a proposition over the
set of propositional variables $V$'', and we keep it in the back of our minds.
\end{node}

\begin{node}\label{prop-000G}%
We can encode the syntax of propositional logic using \SML\ as a
recursive data structure. This reflects the tree-like structure to the
grammar of well-formed formulas \zref{prop-000I}.

If we were to do this in production, we would probably place this in a
structure with the various other utility functions. For example, the
datatype encoding variables might want to include metadata.

\begin{sml}
type Var = string;

datatype Prop = P of Var;

datatype 'a Formula = True
                    | False
                    | Atom of 'a
                    | Not of 'a Formula
                    | And of ('a Formula)*('a Formula)
                    | Or of ('a Formula)*('a Formula)
                    | Implies of ('a Formula)*('a Formula)
                    | Iff of ('a Formula)*('a Formula)
                    | Forall of Var*('a Formula)
                    | Exists of Var*('a Formula);
(* We will work with `Prop Formula` in propositional logic *)
\end{sml}
\end{node}
\end{definition}

\begin{node}[Precedence]\label{prop-000D}%
We should note that logical connectives have the following binding
strength (from weakest to strongest): $\neg$, $\land$, $\lor$,
$\implies$, $\iff$. These are ``left associative''. We may insert
parentheses to alter it. Some examples:
\begin{enumerate}
\item $A\iff (\neg B)\lor C\implies D$ is parsed as $A\iff(((\neg B)\lor C)\implies D)$.
\item $A\implies B\implies C$ is parsed as $(A\implies B)\implies C$.
\end{enumerate}
\end{node}


\begin{node}\label{prop:syntax-0000}%
The set of propositional variables will be useful when trying to
determine if we have a tautology or contradiction or whatever.

\begin{node}\label{prop:syntax-0002}%
Testing for membership should be straightforward.

\begin{sml}
(* in structure Prop *)
fun member _ [] = false
 |  member (p as (P y)) ((P x)::xs) = (x = y) orelse member p xs;
\end{sml}
\end{node}

\begin{node}[Union of sets of propositional variables]\label{prop:syntax-0003}%
We will encode sets of propositional variables as lists without
duplicates, so taking the union of two such finite sets of variables
requires testing for duplicates before merging in.

\begin{sml}
(* merge : ('a -> 'a list -> bool) -> 'a list -> 'a list -> 'a list *)
fun merge member [] rs = rs
 |  merge member ls [] = ls
 |  merge member (h::tl) xs = if member h xs
                              then merge tl xs
                              else merge tl (h::xs)
\end{sml}
\end{node}
\begin{node}\label{prop:syntax-0001}%
Now we can obtain the set of propositional variables:

\begin{sml}
(* in structure Prop *)
(* vars : Prop Formula -> Prop list *)
fun vars False = []
 |  vars True = []
 |  vars (Atom p) = [p]
 |  vars (Not fm) = vars fm
 |  vars (Or (A,B)) = merge member (vars A) (vars B)
 |  vars (And (A,B)) = merge member (vars A) (vars B)
 |  vars (Implies (A,B)) = merge member (vars A) (vars B)
 |  vars (Iff (A,B)) = merge member (vars A) (vars B)
 |  vars _ = [];
\end{sml}
\end{node}
\end{node}
\end{node}
