% 26
\begin{node}\label{nbg-0000}%
We can formalize a minimal notion of ``proper classes'' and treat
sets as a synonym for classes contained in another class.
\end{node}

\begin{puzzle}[\NBG\ in Mizar language]\label{nbg-0001}%
Can we formalize \NBG\ set theory using the Mizar language? Arguably, the
unique primitive notion would be a ``class'' instead of a ``set'', and
we could define ``\texttt{mode set} \verb|->| \texttt{class means ex C
  being class st it in C}''. The primitive notion ``\texttt{in}'' is a
predicate taking a mathematical object as its left argument and a class
as its right argument. We also have a primitive equality predicate.

\begin{node}[References]\label{nbg-0002}%
There are few good references on \NBG\ set
theory. Banakh~\cite{banakh2023classical} has some lecture
notes. Someone has done some heroic work on the English wikipedia page
for \NBG\ set theory. Most texts discussing the \NBG\ axioms follow
G\"{o}del's treatment~\cite{godel1940consistency}.
%% Pulgar\'{\i}n and Uribe-Zapata~\cite{pulgarin2023came} wrote a paper
%% trying to make finitistic metatheory rigorous.
\end{node}

\begin{node}[NBG is about sets]\label{nbg-0003}%
Remember, the whole point of \NBG\ set theory is to reason about sets. We
just get the ability to explicitly make statements concerning proper
classes as a side effect.
\end{node}
\end{puzzle}

\begin{node}[Conventions]\label{nbg-000F}%
We will use lowercase variables $x$, $y$, $z$, \dots, for sets;
uppercase variables $X$, $Y$, $Z$, \dots, for classes; and bold
uppercase variables $\properclass{X}$, $\properclass{Y}$, \dots, for proper classes.
G\"{o}del is credited with inventing this convention.
\end{node}

\begin{definition}\label{nbg-0007}%
Let $x$ and $y$ be mathematical objects. We define the term $\langle x,y\rangle$
to be the set equal to $\langle x,y\rangle := \{\{x,y\}, \{x\}\}$.
\end{definition}

\begin{definition}\label{nbg-000A}%
Let $X$ be a set. We define the term $\dom(X)$ (called the
\define{Domain} of $X$) to be the class equal to
$\dom(X):=\{x\mid\exists y,\langle x,y\rangle\in X\}$ where $x$ and $y$
are variables ranging over sets.
\end{definition}

\begin{node}[Notation]\label{nbg-000B}%
We will denote by $\properclass{U}$ the proper class of the universe of sets.
Also, we will denote by $\properclass{E}$ the proper class of the
membership relation.
In particular, $\dom(\properclass{E})=\properclass{U}$.
\end{node}

\begin{node}[Axioms]\label{nbg-0004}%
The 14 axioms of \NBG\ set theory have the advantage that it is finitely
axiomatizable. This lends itself to working with a metatheory which is
finitistic and recursively enumerable.

\begin{node}[Axiom of extensionality]\label{nbg-0005}%
For all classes $A$ and $B$ such that every object $x$ satisfies $x\in A$
iff $x\in B$, it holds that $A=B$.
\end{node}

\begin{node}[Axiom of pair]\label{nbg-0006}%
For all sets $x$ and $y$ there exists a set $z$ such that for all sets
$u$, $u\in z$ if and only if $u=x$ or $u=y$.
\end{node}

\begin{node}[Axiom of membership]\label{nbg-0008}%
There exists a [proper] class $\properclass{E}$ such that for all sets $z$,
$z\in\properclass{E}$ if and only if there exists sets $x$ and $y$ such
that $x\in y$ and $z=\langle x,y\rangle$.
\end{node}

\begin{node}[Axiom of domain]\label{nbg-0009}%
For any class $X$ there exists a class $D$ such that for all sets $x$,
$x\in D$ if and only if there exists a set $y$ such that $\langle x,y\rangle\in X$.
\end{node}

\begin{node}[Axiom of difference]\label{nbg-000G}%
For any classes $X$ and $Y$, there exists a class $Z$ such that for all
sets $u$ we have $u\in Z$ if and only if $u\in X$ and $u\notin Y$.
\end{node}

\begin{node}[Axiom of product]\label{nbg-000H}%
For any classes $X$ and $Y$ there exists a class $Z$ such that every set
$z$ satisfies $z\in Z$ if and only if there exists sets $x\in X$ and
$y\in Y$ such that $z=\langle x,y\rangle$.
\end{node}

\begin{node}[Axiom of inversion]\label{nbg-000C}%
For any class $X$ there exists a class $Y$ such that for all sets $p$ we
have $p\in Y$ if and only if there exist sets $x$ and $y$ such that
$\langle x,y\rangle\in X$ and $p=\langle y,x\rangle$.
\end{node}

\begin{node}[Axiom of cycle]\label{nbg-000I}%
For all classes $X$ there exists a class $Y$ such that every set $t$
satisfies $t\in Y$ if and only if there exists sets $x$, $y$, $z$ such
that $\langle x,y,z\rangle\in X$ and $t=\langle z,x,y\rangle$.
\end{node}

\begin{node}[Axiom of replacement]\label{nbg-000J}%
For all functions $F$ and sets $x$, the class $F(x)$ is a set.
\end{node}

\begin{node}[Axiom of Union]\label{nbg-000K}%
For any set $x$, the class $\union x=\{z\mid\exists y\in x,z\in y\}$
is a set.
\end{node}

\begin{node}[Axiom of powerset]\label{nbg-000L}%
For any set $x$, the class $\powerset(x)=\{y\mid y\subset x\}$ is a set.
\end{node}

\begin{node}[Axiom of infinity]\label{nbg-000M}%
There exists a set $x\in\properclass{U}$ such that $\emptyset\in x$
and for all sets $n$, if $n\in x$ then $n\cup\{n\}\in x$.
\end{node}

\begin{node}[Axiom of foundation]\label{nbg-000N}%
For all sets $x\in\properclass{U}$, if $x\neq\emptyset$, then there exists a
set $y\in x$ such that every set $z\in y$ satisfies $z\notin x$.
\end{node}

\begin{node}[Axiom of global choice]\label{nbg-000O}%
There exists a class $F$ such that $F$ is a function and for every set
$x\in\properclass{U}$, if $x\neq\emptyset$, then there exists a set $y\in x$
such that $\langle x,y\rangle\in F$.
\end{node}
\end{node}

\begin{node}[Jech's axiomatization]\label{nbg-000D}%
Jech offers, at the end of chapter 6 of~\cite{jech2003set},
the following axiomatization of \NBG\ set theory:
\begin{enumerate}
  \item \begin{enumerate}
  \item Extensionality: $\forall u(u\in X\iff u\in Y)\implies X=Y$
  \item If $X\in Y$, then $X$ is a set
  \item Pairing: for any sets $x$ and $y$, $\{x,y\}$ is a set
  \end{enumerate}
  \item Comprehension $\forall X_{1}\dots\forall X_{n}\exists Y, Y=\{x\mid\varphi(x,X_{1},\dots,X_{n})\}$
    where $\varphi$ is a formula where only set variables are quantified.
  \item \begin{enumerate}
  \item Infinity: there exists an infinite set
  \item Union: for every set $x$, the set $\bigcup x$ exists
  \item Power set: for every set $x$ the powerset $\powerset(x)$ exists
  \item Replacement: If a class $F$ is a function and $x$ is a set, then
    $\{F(z)\mid z\in x\}$ is a set
  \end{enumerate}
  \item Regularity
  \item Choice: There is a function $F$ such that $F(x)\in x$ for every
    nonempty set $x$.
\end{enumerate}
\end{node}

\begin{node}[In proof assistants]\label{nbg-000E}%
When it comes to the benefits of finite axiomatizability,
Hester~\cite{hester2019automated} investigated whether there's any
benefits. It appears whatever benefits we might have gotten are lost in
the naive implementation, since we need to include extra predicates for
quantifiers over sets.
\end{node}

\begin{node}[Uselessness in Category Theory]\label{nbg-000P}%
We might think that \NBG\ set theory would be useful when formalizing
category theory. And we would be wrong. If $\cat{C}$ is a large
category, then we cannot prove the $n$-fold product category
$\cat{C}^{n}$ exists using \NBG\ set theory. If $\cat{C}$ and $\cat{D}$
are large categories, we cannot define the category of functors
$\cat{D}^{\cat{C}}$ even using \MK\ axioms. This usually leads us to
Tarski--Grothendieck set theory.
\end{node}
